Monitoring forest health using hyperspectral imagery: Does feature selection improve the performance of machine-learning techniques?

### 2020-04-30 docs/00-manuscripts/ieee/ieee-manuscript.tex
upd results and discussion for lasso scores

\usepackage{booktabs}
\usepackage{adjustbox}

@@

% - RMSE around 27
% - MBO tuned penalized methods are substantially worse
\noindent Overall, the response variable 'tree defoliation' could be modeled with an \ac{RMSE} of 28 \% percentage points (p.p.).
SVM showed no differences in RMSE across feature sets whereas other learners (RF, SVM, XGBoost, lasso and ridge) differed up to seven percentage points (\autoref{fig:perf-result}).
Ridge and lasso faced major issues up to the point of not achieving an error below 50 for datasets HR-NRI-VI and NRI-VI.
SVM showed the best performance of all learners with a difference of around three percentage points to the next best model (RF) (\autoref{tab:best-learner-perf}).
Performance differences between test folds were large: Predicting on Luiando resulted in an RMSE of 9.0 for learner SVM (without filter) but up to 54.26 p.p. when testing on Laukiz2 (\autoref{tab:svm-single-fold-perf}).

@@

% learner
% - SVM Car is the overall winner
% - Ridge-CV hit and miss (hit = NRI, miss = VI)
% - all MBO penalized methods have no difference across tasks
SVM combined with the 'Carscore' filter achieved the best overall performance (RMSE of 27.98 p.p.) (\autoref{tab:perf-top-10}).
%L1 penalized regression showed slightly better results in three out of six tasks when being optimized using the package internal 10-fold CV compared to MBO (RMSE of 50 vs 58).
Regression with ridge penalty (L2) showed a high variance when comparing results across tasks: In two out of six tasks (all including VI variables) the error was out of bounds (HR-NRI-VI, HR-VI, VI). For HR-NRI and HR RMSE was around 40 p.p., and for task NRI the ridge learner achieved almost the same score as the best performing model (32.9 p.p) (\autoref{tab:perf-top-10}).
In all settings for which ridge showed such a high error, only one observation in one fold was predicted which such a high value (in the millions).
The influential observation showed no obvious signs of being anomalous, e.g. extreme values in individual features or in principal components of features.
This then caused the error estimates of these folds and the average estimate across all folds to be out of bounds.

FIXME: Mention: All penalized methods tuned using MBO showed exactly the same error across all tasks.
FIXME: Mention the bad perf of RIDGE on VI

@@

@@

% latex table worst 10 absolute performances
%\input{performance-worst-10}

@@

@@

We were a bit surprised about the performances of the penalized methods lasso and ridge.
The former was not able to achieve an RMSE below 46.0 p.p. across any feature set while ridge came at least somewhat close to RF and XGBoost for some feature sets.
However, there were also three settings (VI, NRI-VI, HR-NRI-VI) for which the RMSE of the ridge model was unreasonably large.
Careful inspection of the hyperparameter optimization of these learners did not reveal any implementation mistakes. However, one observation turned out to be influential even though careful multivariate exploratory data analysis showed no reason for flagging this observation a priori.
We therefore left it in the model, but we acknowledge that these learners seemed particularly sensitive to the input data.

FIXME: Talk about L1 and L2 model performance

@@

@@

% write about hyperspectral remote sensing
The field of (hyperspectral) remote sensing has a strong focus on using random forest for modeling in recent years \cite{belgiu2016}.
In high-dimensional scenarios, tuning parameter \texttt{\(m_{try}\)} becomes computational expensive.
To account for this and the high-dimensionality in general, studies used feature selection approaches like semi-supervised feature extraction \cite{xia2015}, wrapper methods \cite{fassnacht2014, feng2016}, PCA and adjusted feature selection \cite{rochac2016}.
We could not find a study which made use of filter methods in combination with hyperparameter tuning in the field of (hyperspectral) remote sensing.
Potential reasons could be an easier access of applying wrappers methods and a higher general awareness of such compared to filter methods.
Applying the filter-based feature selection methodology shown in this study and its related code provided in the research compendium might be a helpful reference for upcoming modeling studies using hyperspectral remote sensing data.

% RS studies with more/other algs than RF
Studies which include more than one model usually operate in a low-dimensional setting 
\cite{xu2019}.

>>>>

\usepackage{booktabs}
\usepackage{multirow}
\usepackage{adjustbox}

@@

% - RMSE around 27
% - MBO tuned penalized methods are substantially worse
\noindent Overall, the response variable 'tree defoliation' could be modeled with an \ac{RMSE} of 28 \% percentage points (p.p.).
SVM showed no differences in RMSE across feature sets whereas other learners (RF, SVM, XGBoost, lasso and ridge) differed up to seven percentage points (\autoref{fig:perf-result}).
Ridge faced major issues in four tasks due to one observation that was predicted off the scale.
SVM showed the best performance of all learners with a difference of around three percentage points to the next best model (RF) (\autoref{tab:best-learner-perf}).
Performance differences between test folds were large: Predicting on Luiando resulted in an RMSE of 9.0 for learner SVM (without filter) but up to 54.26 p.p. when testing on Laukiz2 (\autoref{tab:svm-single-fold-perf}).

@@

% learner
% - SVM Car is the overall winner
% - Ridge-CV hit and miss (hit = NRI, miss = VI)
% - all MBO penalized methods have no difference across tasks
SVM combined with the 'Carscore' filter achieved the best overall performance (RMSE of 27.98 p.p.) (\autoref{tab:perf-top-10}).
%L1 penalized regression showed slightly better results in three out of six tasks when being optimized using the package internal 10-fold CV compared to MBO (RMSE of 50 vs 58).
Regression with ridge penalty (L2) showed a high variance when comparing results across tasks: In four out of six tasks (all which VI variables and HR-NRI) the error was out of bounds (\autoref{tab:perf-worst-10}).
For NRI and HR RMSE was 31.16 and 35.45 p.p., respectively.
In all settings for which ridge showed such a high error, only one observation in one fold was predicted which such a high value (in the millions).
The specific observation showed no obvious signs of being anomalous, e.g. extreme values in individual features or in principal components of features.
This one outlier caused the error estimates of these folds and the average estimate across all folds to be off the scale.

@@

% Selected percentage of features across plots
High differences were observed between the amount of features selected during tuning for the subsequent fitting process.
Most features were selected during optimization for plot Laukiz2 and least for Laukiz1 (\autoref{tab:tune-perc-sel-features}).
RF used only one feature for plots Luiando and Oiartzun (7675 $\times$ 0.00004) while for Laukiz1 34 and for Laukiz2 230 features were seen as optimal by the optimizer.
In contrast, XGBoost and SVM used in all cases but Laukiz1 (less than 50 features) more than two-third of all available features.

@@

% latex table worst 10 absolute performances
\input{performance-worst-10}

@@

% latex table percentage of selected features during tuning
\input{tune-perc-sel-features.tex}

@@

Penalized methods showed promising performances, especially when taking runtime into account.
When removing features with a correlation of 1, lasso is able to score performances around 31 p.p. and shows somewhat surprisingly the best performance on the VI task.
The same general conclusion applies to ridge if one discards the problematic results on four tasks.
A careful inspection of the hyperparameter optimization procedure of these models did not reveal any implementation mistakes.
One observation turned out to be influential in causing the outlier prediction value even though careful multivariate exploratory data analysis showed no reason for flagging this observation a priori.
We therefore left it in the model, but we acknowledge that these learners seemed particularly sensitive to the input data.

@@

The high differences of selected features per plot plot during tuning give interesting insights into internals of the used models (\autoref{tab:tune-perc-sel-features}).
While SVM and XGBoost require a substantial portion of features in most cases to achieve robust predictions, RF is able to achieve the best results with a relatively low amount of features.
This fact leads to reduced computational runtime, especially if model parameters are dependent on the amount of features (e.g. \texttt{\(m_{try}\)}).
Hence, regardless of the potential advantage of using filters for increased predictive performance, these can have a strong positive effect on runtime, at least for RF in this study.
However, the results of (\autoref{tab:tune-perc-sel-features}) should be taken with care as they rely on single model-filter combinations and are subject to random variation.
More in-depth research is needed to investigate the effect of filters on other criteria (such as runtime) than performance, leading to a multi-criteria optimization problem.

@@

% write about hyperspectral remote sensing
The field of (hyperspectral) remote sensing has a strong focus on using random forest for modeling in recent years \cite{belgiu2016}.
In high-dimensional scenarios, tuning parameter \texttt{\(m_{try}\)} becomes computational expensive.
To account for this and the high-dimensionality in general, studies used feature selection approaches like semi-supervised feature extraction \cite{xia2015}, wrapper methods \cite{fassnacht2014, feng2016, georganos2018}, PCA and adjusted feature selection \cite{rochac2016}.
We could not find a study which made use of filter methods in combination with hyperparameter tuning in the field of (hyperspectral) remote sensing.
Potential reasons could be an easier access of applying wrappers methods and a higher general awareness of such compared to filter methods.
Applying the filter-based feature selection methodology shown in this study and its related code provided in the research compendium might be a helpful reference for upcoming modeling studies using hyperspectral remote sensing data.

% RS studies with more/other algs than RF	
Remote sensing studies comparing multiple models often operate in a low-dimensional predictor space \cite{xu2019} or use wrapper methods explicitly \cite{georganos2018}.
